Cataclysmic variables are a key population of gravitational wave sources for LISA

### 2023-06-15 referee_response.txt

We appreciate the referee's careful study of our paper and respond to the comments and questions below inline. Changes to the manuscript are highlighted in bold and referenced by paragraph in the referee response. We hope the referee finds this newly submitted manuscript publishable by MNRAS Letters.

Abstract: I would define what a CV is in here and also include the value of the minimum orbital period that is mentioned.
 --> We have added two clarifications, one in sentence 2 and one in sentence 8.

Introduction: 
In the third paragraph, it would be useful to include the LISA frequency range corresponding to the orbital period range.
--> We have included the LISA frequency range corresponding to the orbital period range in paragraph 3.

Beginning p2: rho_CV is not defined.
--> We have now clarified that rho_CV and rho_AMCvn are space densities.

3rd paragraph on p2, about the reason for the period gap: This paragraph is hard to follow, with a lot of jargon (the mass-radius index, an adiabatic value, a stellar vs non-stellar object…).  I would rephrase and/or add more context.
--> We have added a definition of the mass-radius index, clarified that adiabatic mass loss is faster than the radius can respond, and clarified sub-stellar.

Section 2:
Second paragraph:
-I believe something is missing in the first sentence.
--> We have reworded the first sentence.
-How was equation 1 obtained?
--> This equation was obtained empirically such that the period gap limits are enforced during our sampling procedure.

Section 3:
Second to last paragraph: I’m not sure if the considered LDC population does or does not contain interacting binaries given the sentence “For this study, we use the same Galactic population as was used for the LDC2a-v1 data set provided by the LISA Data Chal- lenges (LDCs) (Le Jeune & Babak 2022) which consisted of only de-tached double WD binaries”  and then just after « one which contains detached and accreting double WD binaries from the LDC “.  This seems to be a contradiction.
*** WE STILL NEED TO CLARIFY THIS **

Section 4:
-       ASD is undefined, I think.
-       Fig2: It would be useful to have the period on the top axis.
            --> We have added the orbital period on the top axis.
-       Fig2: I guess this figure does not include DWD confusion noise?
            --> This is correct; we have clarified this in the caption.
-       End of 2nd paragraph on p4: it’s-> its
            --> We appreciate the catch from the referee and have fixed the typo. 
-       Same sentence: I am not sure I follow the argument here: are you expecting most of the science to be coming from the foreground because the absence of frequency derivative implies lack of chirp mass/distance measurement, which limits the identification of the objects?
    --> This is exactly the reasoning; we have added a clarifying sentence.

-       Second paragraph on p5: I am not sure I understand the implication here: The sources will stop interacting, but they are still emitting GWs. Does this mean that effectively there will be a continuum and without EM counterpart we cannot make any distinction? The measurable feature will only be a pile-up around 80 min? I agree that interacting and non-interaction populations have slightly different evolutions, but given the small population, can we expect this effect to be measurable?
    -->

### 2023-06-15 src/tex/ms.tex
updates for referee response;

% Abstract of the paper
\begin{abstract}
AM~CVn stars are well established as an important population of gravitational wave (GW) sources for the planned \lisa\ mission. The less compact – but more numerous – cataclysmic variables (CVs) have not been studied carefully in this context. Here, we estimate the predicted GW signals from the Galactic population of cataclysmic variables and evaluate their significance for \lisa. First, we find that at least three known systems are expected to produce strong enough signals to be individually resolved within the first four years of \lisa's operation. Second, cataclysmic variables will contribute significantly to the \lisa\ Galactic binary background, thus limiting the mission's sensitivity in the relevant frequency band. Third, we predict a spike in the unresolved GW background at a frequency corresponding to the cataclysmic variable minimum orbital period. This excess noise may impact the detection of other low signal-to-noise systems near this characteristic frequency. Fourth, we note that the amplitude and location of the excess noise spike associated with $P_{\rm{min}}$ can be used to measure the cataclysmic variable space density and period bounce location with complementary and simple GW biases compared to the biases and selection effects that plague samples selected from electromagnetic signals. Our results highlight the need to explicitly include the Galactic cataclysmic variable population in \lisa\ mission planning, both as individual GW sources and generators of background noise, as well as the exciting prospect of characterising the cataclysmic variable population through their GW emission.
\end{abstract}

@@

Given the effort that has been expended on predicting the GW signals produced by AM~CVn stars and detached double degenerates -- both individually and as a population -- it is somewhat surprising that their close cousins, cataclysmic variables (CVs), have not received much attention in this context. CVs are compact binary systems in which a WD accretes material from a hydrogen-rich main-sequence or sub-stellar donor star that fills its Roche lobe. Their orbital periods lie in the range $75~\mathrm{min} \lesssim P_{\rm{orb}} \lesssim 1~\mathrm{day}$, part of which still falls into the \lisa\ band. Perhaps more importantly, the space density of CVs is much higher than that of AM~CVn stars. The best current estimates are $\rho_{CV} = 4.8^{+0.8}_{-0.6} \times 10^{-6}~\mathrm{pc^{-3}}$\citep{pala20} and $\rho_{AM} = 5\pm 3 \times 10^{-7}~\mathrm{pc^{-3}}$\citep{carter13} for CVs and AM~CVn systems respectively. This will cause an elevated background of signals in the relevant frequency range.

@@

%The period minimum is also associated with a change in the structure of the %donor star. 
Below the period gap, the thermal time-scale of the donor increases significantly faster than the mass-loss time-scale, causing the donor's mass-radius index to evolve from a near-thermal equilibrium value of  $\zeta_{th} \simeq 1$ towards the adiabatic value of $\zeta_{ad} = -1/3$. However, the geometry of a semi-detached binary system implies that the donor must also satisfy an orbital period - density relation. It is easy to show from this that the orbital period must start to increase again once the mass-radius index of the donor reaches $\zeta = 1/3$ \citep[see e.g.][]{knigge11}. In practice, this happens at $P_{\rm{min}} \simeq 80$~min and roughly coincides with the transition of the donor from a stellar to a sub-stellar object. The period minimum is expected -- and observed -- to coincide with a pile-up in the CV period distribution, the so-called "period spike" \citep[][]{gansicke09}.

@@

\cite{pala20} presents the first volume-limited sample of CVs selected using parallax measurements provided by \gaia. The sample consists of $42$ CVs within $150\,\rm{pc}$ and is estimated by them to be $\approx 75\%$ complete. Fig.~\ref{fig:porb} shows the orbital period distribution for these 42 systems in both cumulative (dotted orange curve) and differential forms (thick orange curve).

@@

It is also interesting to note that there may be the possibility of detecting other CV systems through their GW emission which have been missed due to electromagnetic detection biases and not included in the current $150~\rm{pc}$ volume-limited sample. In our simulated $150~\rm{pc}$ population, which represents a single statistical realization, we find $12$ CVs in addition to the $42$ systems included in the \citet{pala20} catalog. Of those $12$, the CV with the largest ASD has a distance of $65~\rm{pc}$ and a GW frequency $f_{\rm{GW}} = 0.26~\rm{mHz}$. This system is likely an outlier, but is consistent with our adopted space density and orbital period distribution. Perhaps more important are the CVs with distances beyond $100~\rm{pc}$ that have orbital periods near $P_{\rm{min}}$, thus aiding in their GW detection while having dimmer EM signatures due to the low mass of the donor star. This population could aid in providing tighter constraints on the local space density of the CV population. Finally, we note that it is almost certainly impossible for LISA to observe any orbital evolution for any of the individually resolved CVs based on the models of \citet{knigge11}. This is in stark contrast to AM CVn systems \citep[e.g.][]{Nelemans2004, Kremer2017, Breivik2018} and suggests that the bulk of the scientific return for the CV population will come from observations of it's contribution to the LISA confusion foreground.

\begin{figure}
	\includegraphics[width=0.5\textwidth]{figures/fig2.pdf}
    \caption{Amplitude Spectral Density (ASD) computed assuming four years of \lisa\ operations as a function of GW frequency. The corresponding \lisa\ instrument noise is shown with the dotted black line. The $150~\rm{pc}$ \citet{pala20} sample is shown by the large orange circles, while the remaining simulated systems within $150~\rm{pc}$ are shown in orange. Blue dots show the simulated $1~\rm{kpc}$ CV population using the inferred \citet{pala20} space density.}
    \label{fig:asd}
    \script{fig2.py}
\end{figure}

@@

The excess GW noise produced by CVs is directly related their high space density and exhibits a specific form. As CVs evolve towards short orbital periods, they "pile-up" towards the so called orbital period minimum near $P_{\rm{min}} \simeq 80$~min. As a result, the GW noise excess produced by CVs increases towards the corresponding GW frequency, $f_{max} = 2/P_{\rm{min}} \simeq 0.4$~mH, but then drops sharply for frequencies above this limit. The distinctive shape of this feature should allow its detection and characterization among other sources of GW noise. Its amplitude will then immediately allow a new and robust estimate of the Galactic CV space density, without the biases that affect all electromagnetically selected CV samples.

>>>>

% Abstract of the paper
\begin{abstract}
AM~CVn stars are well established as an important population of gravitational wave (GW) sources for the planned \lisa\ mission. The less compact – but more numerous – cataclysmic variables (CVs)\textbf{, binaries containining white dwarfs which are observed to accrete hydrogen-rich material from a companion,} have not been studied carefully in this context. Here, we estimate the predicted GW signals from the Galactic population of cataclysmic variables and evaluate their significance for \lisa. First, we find that at least three known systems are expected to produce strong enough signals to be individually resolved within the first four years of \lisa's operation. Second, cataclysmic variables will contribute significantly to the \lisa\ Galactic binary background, thus limiting the mission's sensitivity in the relevant frequency band. Third, we predict a spike in the unresolved GW background at a frequency corresponding to the cataclysmic variable minimum orbital period. This excess noise may impact the detection of other low signal-to-noise systems near this characteristic frequency. Fourth, we note that the amplitude and location of the excess noise spike associated with \textbf{$P_{\rm{min}}\sim80~\mathrm{min}$} can be used to measure the cataclysmic variable space density and period bounce location with complementary and simple GW biases compared to the biases and selection effects that plague samples selected from electromagnetic signals. Our results highlight the need to explicitly include the Galactic cataclysmic variable population in \lisa\ mission planning, both as individual GW sources and generators of background noise, as well as the exciting prospect of characterising the cataclysmic variable population through their GW emission.
\end{abstract}

@@

Given the effort that has been expended on predicting the GW signals produced by AM~CVn stars and detached double degenerates -- both individually and as a population -- it is somewhat surprising that their close cousins, cataclysmic variables (CVs), have not received much attention in this context. CVs are compact binary systems in which a WD accretes material from a hydrogen-rich main-sequence or sub-stellar donor star that fills its Roche lobe. Their orbital periods lie in the range $75~\mathrm{min} \lesssim P_{\rm{orb}} \lesssim 1~\mathrm{day}$ {\bf{($0.02~\mathrm{mHz}\lesssim f_{\mathrm{GW}}\lesssim 0.44~\mathrm{mHz}$)}}, part of which still falls into the \lisa\ band. Perhaps more importantly, the space density of CVs is much higher than that of AM~CVn stars. The best current \textbf{space-density} estimates are $\rho_{CV} = 4.8^{+0.8}_{-0.6} \times 10^{-6}~\mathrm{pc^{-3}}$\citep{pala20} and $\rho_{AM} = 5\pm 3 \times 10^{-7}~\mathrm{pc^{-3}}$\citep{carter13} for CVs and AM~CVn systems respectively. This will cause an elevated background of signals in the relevant frequency range.

@@

%The period minimum is also associated with a change in the structure of the %donor star. 
Below the period gap, the thermal time-scale of the donor increases significantly faster than the mass-loss time-scale. \textbf{This causes the donor's mass-radius index, $\mathrm{d}\ln R/\mathrm{d}\ln M$, to evolve from a near-thermal equilibrium value of $\zeta_{th} \simeq 1$ towards the adiabatic value of $\zeta_{ad} = -1/3$, where mass loss occurs at faster rate than the radius can respond}. However, the geometry of a semi-detached binary system implies that the donor must also satisfy an orbital period - density relation. It is easy to show from this that the orbital period must start to increase again once the mass-radius index of the donor reaches $\zeta = 1/3$ \citep[see e.g.][]{knigge11}. In practice, this happens at $P_{\rm{min}} \simeq 80$~min and roughly coincides with the transition of the donor from a stellar to a sub-stellar object \textbf{which is no longer fusing in its core}. The period minimum is expected -- and observed -- to coincide with a pile-up in the CV period distribution, the so-called "period spike" \citep[][]{gansicke09}.

@@

\textbf{The first volume-limited of CVs was selected using parallax measurements provided by \gaia was released by \cite{pala20}.} The sample consists of $42$ CVs within $150\,\rm{pc}$ and is estimated by them to be $\approx 75\%$ complete. Fig.~\ref{fig:porb} shows the orbital period distribution for these 42 systems in both cumulative (dotted orange curve) and differential forms (thick orange curve).

@@

It is also interesting to note that there may be the possibility of detecting other CV systems through their GW emission which have been missed due to electromagnetic detection biases and not included in the current $150~\rm{pc}$ volume-limited sample. In our simulated $150~\rm{pc}$ population, which represents a single statistical realization, we find $12$ CVs in addition to the $42$ systems included in the \citet{pala20} catalog. Of those $12$, the CV with the largest ASD has a distance of $65~\rm{pc}$ and a GW frequency $f_{\rm{GW}} = 0.26~\rm{mHz}$. This system is likely an outlier, but is consistent with our adopted space density and orbital period distribution. Perhaps more important are the CVs with distances beyond $100~\rm{pc}$ that have orbital periods near $P_{\rm{min}}$, thus aiding in their GW detection while having dimmer EM signatures due to the low mass of the donor star. This population could aid in providing tighter constraints on the local space density of the CV population. Finally, we note that it is almost certainly impossible for LISA to observe any orbital evolution for any of the individually resolved CVs based on the models of \citet{knigge11}. \textbf{A lack of an observed orbital evolution will also limit a distance measurement or precise sky position.} This is in stark contrast to AM CVn systems \citep[e.g.][]{Nelemans2004, Kremer2017, Breivik2018} and suggests that the bulk of the scientific return for the CV population will come from observations of its contribution to the LISA confusion foreground.

\begin{figure}
	\includegraphics[width=0.5\textwidth]{figures/fig2.pdf}
    \caption{Amplitude Spectral Density (ASD) computed assuming four years of \lisa\ operations as a function of GW frequency. The corresponding \lisa\ instrument noise, \textbf{without the Galactic foreground,} is shown with the dotted black line. The $150~\rm{pc}$ \citet{pala20} sample is shown by the large orange circles, while the remaining simulated systems within $150~\rm{pc}$ are shown in orange. Blue dots show the simulated $1~\rm{kpc}$ CV population using the inferred \citet{pala20} space density.}
    \label{fig:asd}
    \script{fig2.py}
\end{figure}

@@

The excess GW noise produced by CVs is directly related their high space density and exhibits a specific form. As CVs evolve towards short orbital periods, they "pile-up" towards the so called orbital period minimum near $P_{\rm{min}} \simeq 80$~min. As a result, the GW noise excess produced by CVs increases towards the corresponding GW frequency, $f_{\rm{max}} = 2/P_{\rm{min}} \simeq 0.4$~mH, but then drops sharply for frequencies above this limit. The distinctive shape of this feature should allow its detection and characterization among other sources of GW noise. Its amplitude will then immediately allow a new and robust estimate of the Galactic CV space density, without the biases that affect all electromagnetically selected CV samples.
